---
title: "Data preparation and exploration"
description: |
 Here we joined our datasets and explored through visualization.
author:
  - name: Asha
    url:
    affiliation: University of Oregon
    affiliation_url:
date: 11-28-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidymodels)
library(baguette)
library(future)
library(here)
library(rio)
library(corrplot)
library(skimr)

```
## 1. Packages 

For the purpose of data loading and cleaning, we are using following packages in R:
`{tidyverse}`,
`{here}`,
`{rio}`, and
`{skimr}`


## 2. Joining the datasets

For the purpose of demonstration, we will be using 1% of the data with `sample_frac()` to keep computing time low. All our datasets have `school ids` which we used as `key` to join the datasets.

After loading our three datasets, we joined them together to make one cohesive dataset, to be used for ML modeling. After joining, the dataset contains student-level variables (e.g. gender, ethnicity, enrollement in special education, etc.) as well as district-level variables ( school longitude and latitude, proportion of free and reduced lunch, etc.). All of these variables will be used in our ML models to predict student score in the statewide assessment. Here is the preview of our final dataset, ready to be used for ML modeling.



```{r, echo=FALSE}
set.seed(3000)

full_train <- import(here("data", "train.csv"), setclass = "tbl_df") %>%
  select(-classification) %>%
  mutate_if(is.character, factor) %>%
  mutate(ncessch = as.numeric(ncessch))

bonus <- import(here("data", "bonus_data.csv")) %>%
  mutate(pupil_tch_ratio = as.numeric(pupil_tch_ratio)) %>%
  mutate(ncessch = as.double(ncessch))

## joining data
data <- left_join(full_train, bonus) %>%
  sample_frac(0.01)

skim(data) %>%
  select(-starts_with("numeric.p")) #remove quartile

```


## 3.Data Exploration

Correlation: Figure 1 displays significant (`p < 0.05`) correlation between independent (predictors) numeric variables. `Red dots` show significant pearson's correlation coefficient between 0 to 1 and `white dots` show significant coefficient between -1 to 0. Blank spaces are not significant.

```{r, fig.width=20, fig.height=18, echo=FALSE}

data_corr <- data %>%
  select(-contains("id"), -ncessch) %>% 
  select_if(is.numeric) %>% 
  select(score, everything()) %>%
  cor(use = "complete.obs")

# Function to calculate significant pearson's correlation

cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

p.mat <- cor.mtest(data_corr)

data_corr %>% 
  corrplot::corrplot(method = "circle", type = "lower",
                     col=c("white", "red"),
                     bg="lightblue", p.mat = p.mat, 
                     sig.level = 0.05, insig = "blank",
                     tl.col="black", tl.srt=45,
                     order = "hclust",
                     title = "Figure 1: Corelation between numeric variables",
                     mar=c(0,0,1,0))


```
We looked if pupil-teacher ratio and score differ by county.

```{r, fig.width=10, fig.height=10, echo=FALSE}


ggplot(data, aes(pupil_tch_ratio, score)) +
  geom_point(aes(color = county)) + 
  scale_x_continuous("pupil_tch_ratio") +
  scale_y_continuous("Score") +
  theme_bw() + labs(title="Pupil-Teacher ratio by Score and County in Oregon") +
  facet_wrap(~county) +
  theme(legend.position = "none")

```
We explored adult qualification by economic disadvantage and grade.

```{r, fig.width=10, fig.height=10, echo=FALSE}
library(dplyr)

data_factor1 <- data %>%
  select(county, starts_with("percent_"), econ_dsvntg, 
         enrl_grd) %>%
  pivot_longer(6:14,
               names_to = "Adult_Qualification",
               values_to = "Percentage") %>%
  group_by(county) %>%
  drop_na()


ggplot(data_factor1, aes(reorder(Adult_Qualification, Percentage),
       Percentage)) +
  geom_col(aes(fill = econ_dsvntg), position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, size = 10)) +
  coord_flip() +
  scale_fill_manual(values=c("#CC6666","#66CC99")) +
  labs(title = "Adult qualification by economic disadvantage & grade in Oregon") +
  facet_wrap(~enrl_grd)
  
```




